Exploiting Intrinsic Noise to Achieve Heterogeneous Control of Cell Differentiation Dynamics in the Presence of Time Delays, Model Uncertainties, and Extrinsic Variability

### 2023-07-31 paper.tex
update

While most previous control efforts focussed on deterministic ordinary differential equation (ODE) formulations of the control problem, a few pioneering theoretical works have extended these analysis to include noise and have used stochastic simulations to show how systems can be guided towards unstable points by oscillating around those points \cite{Guarino2020}. 
It has recently been shown that new control techniques that leverage the complete probability distribution information of the system could actually harness the noise of single-cell gene regulation to achieve more complicated control objectives. 
For example, inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that noise could be exploited to achieve independent control of multiple cells using a single input, even despite uncertain parameters or time delays due to maturation of fluorescent proteins or limited observation of the regulatory proteins. 
In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{XXX} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure\ \ref{cartoons}A, top promoter).
We then proposed the addition of a positive auto-regulation (Figure\ \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.

@@

\subsection{Model}\label{sec:Model}
To assess the impact of model approximations on the implementation of noise-enhanced control strategies, we begin with the one-species model proposed by May et al. \cite{May2021} (Fig. \ref{cartoons}A). 
This model comprises two reactions for production and degradation of the key protein. 
The nonlinear, UV-dependent production is defined by the following equation:
\begin{equation}
\nu_1(x,t)= \kappa\frac{x^\eta}{x^\eta+\beta^\eta}+ k_0 + u(UV(t)),
\label{prodRate}
\end{equation}
%
%These model dynamics were then simplified to a single ODE with auto-regulation describing the protein accumulation as: 
%\begin{equation}
%\frac{dp}{dt}=u(x_1,x_2)  + \kappa\frac{p^\eta}{p^\eta+\beta^\eta}+ k_0 -\gamma p,
%\label{rateeq}
%\end{equation}
%
where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches its half maximal strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the leakage rate from both promoters; and $u(UV(t))$ is the UV-dependent strength of the T7 promoter. 
Feedback enables the external modulation of the light input using the state of the system, thereby controlling the T7 promoter strength as a function of state rather than time and eliminating the explicit time dependance in $u(UV(t))$.
Protein degradation is assumed to be a first order process with rate $\gamma$:
\begin{equation}
\nu_2(x) = \gamma x.
\end{equation}
All baseline parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$ are presented in Table 1~\cite{May2021} and are fixed throughout the current study.

@@

Optimization of $\mathbf{u}^{FAC}$ was performed using a gradient descent method to minimize $J$. Since the square root is a monotonically increasing function, minimizing $J^2$ results in the same control as would minimizing $J$ directly. Therefore, we calculate the negative gradient $-\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ and walking in a small step $du$ in that direction. The calculation of the gradient is given by
\begin{equation}
\frac{d(J^2)}{d\mathbf{u}^{FAC}}=C \frac{dP}{d\mathbf{u}^{FAC}}
\end{equation}
Where $\frac{dP}{d\mathbf{u}^{FAC}}$ can be solved using general minimized residual calculation of
\begin{equation}
A \frac{dP}{d\mathbf{u}^{FAC}}=B \mathbf{u}^{FAC}
\end{equation}
to find $\frac{dP}{d\mathbf{u}^{FAC}}$.
Optimization of $\mathbf{u}^{PAC}$ was done by restricting $\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ in ambient space to search a low dimensional subspace with a linear lift operator D. Then the subspace search is given by. 
\begin{equation}
\frac{d(J^2)}{d\mathbf{u}^{PAC}}=C \frac{dP}{d\mathbf{u}^{PAC}} = C D \frac{dP}{d\mathbf{u}^{FAC}}  
\end{equation}
%Root mean squared error was calculated by taking the square root of the score.
%\begin{equation}
%RMSE=\sqrt{J}
%\end{equation}

In \cite{May2021}, we optimized each of these two controllers to find the control laws $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ that minimize $J^2$ for a single target state $\mathbf{T} = [10,30]$, but in this work extend it to any target point in the state space.

@@

Delays are inherent in any realistic control systems, and it this case delays would be expected to arise due to the time needed for various biochemical reactions such as formation of complete polymerases, activation of promoters, transcription and transport of mature mRNA, and the translation and maturation of protein \cite{Cai2007}. 
Additional delays would also arise from data analysis, decision making, and actuation dynamics. 
To investigate the effects of time delay on control actuation, we devised a simple time-delay stochastic simulation algorithm. 
This algorithm records the state history after each stochastic jump, enabling reconstruction of the population history of the species and control input.  
Using this information, the time-delayed control signal at time $t$ can be specified as:

\begin{equation}
u_{\tau}(t)=\left\{
\begin{array}{rl}
      0 ,&\textrm{ for }  t \leq \tau, \\
      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\textrm{ for }   t > \tau,\\
\end{array}\right. 
\label{timeDelaySSA}
\end{equation}
where $\tau$ is the time delay between observation and actuation, and ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$). 
\brian[Explain how you modify the SSA for the delayed reaction SSA.  Look through the literature to cite someone else who has implenented a similar method.]
We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.

@@

\subsection{Dynamic Setpoint Control}\label{sec:Dynamic}
Because optimizing the control law for a static set point as described in Section \ref{sec:Quantification} requires differentiation of the CME (Eq. \ref{XXX}) with respect to the control signal at each state, this calculation is approaching the limits of current feasibility.
Extending these calculations to optimize controllers for a dynamically moving set point is much harder and would likely require intractable numerical simulation or development of new mathematical approaches that are beyond the scope of the current study. 
To circumvent this challenge, we instead propose a simple alternative to sweep though a set of controllers over time which target different static points in piecewise fashion to mimic a dynamically moving path.

 We start by describing a discrete path  of target points in state space which cycle at fixed frequency.
  Then, we optimize controllers for each target point and cycling through the controllers over time to create the final controller $\mathbf{R}(\mathbf{x},t)$.
Let $\mathbf{p}(t)$ be a continuous and cyclicly moving point in state space
\begin{equation}
\mathbf{p(t)}=[x_1(t),x_2(t)]
\end{equation}
where $x_1(t)$, and $x_2(t)$ are both cyclic functions of time $\in \mathbf{R}$.
Then, $\mathbf{q}$ is a collection of points $q_i$ which evaluate the cyclic function over its natural period (T) in n steps. 
\begin{equation}
\mathbf{q}=[q_1,\ldots ,q_n]=[\mathbf{p}(t_1),\ldots,\mathbf{p}(t_n)]
\end{equation}
where each $t_i$ comes from linearly spaced points between $0$ and $T$, in $n$ steps 
\begin{equation}
t_i = [t_1,\ldots,t_n]=[0,\frac{1}{n} T,\ldots,\frac{n}{n}T]
\end{equation}
Then, we optimize a set of controllers which target each $q_i$ in the discretized path.
\begin{equation}
\mathbf{U}_i(x)=[\mathbf{u}_1(\mathbf{x}),...,\mathbf{u}_n(\mathbf{x})]
\end{equation}
where $\mathbf{U}_i(x)$

Then, we cycle through each controller at a fixed frequency by using a function which the cycles the controller index ($\phi(t)$) over time. \begin{equation}
\phi(t)=\lceil \frac{n}{T} mod(t,T)\rceil = \lceil n \ f \ mod(t,\frac{1}{f})\rceil
\end{equation}
where $f$ the cycling frequency given by $1/T$. Then the final control equation is given by 
 \begin{equation}
\mathbf{R}(\mathbf{x},t)=\bf{U}_{\phi(t)}(x)
\end{equation}
Where $\mathbf{R}(\mathbf{x},t)$ returns the control input scalar as a function of state and time. 
The dynamical system equation becomes
\begin{equation}
\frac{d}{dt}\mathbf{P}=\mathbf{A}+\mathbf{B}\bf{U}_{\phi(t)}(x)
\end{equation}

\section{Results}

@@

 In this section, we seek to show that noise-exploiting controllers are effective at controlling a noisy system to a dynamically moving target point. To do this, we show that control performance is robust to a range of parameter uncertainties despite not making adjustments to the controler. Then, we extend the control performance analysis to understand how time delay and system scale alters control performance. Then, we find regions in target space which are easiest to control by optimizing individual controllers for each target. We extend these analysis to change the target point over time in a Peicewise fashion. 
 \subsection{Stochastic SIMO Optogenetic Control can Remain Effective Despite Small Parameter Errors or Extrinsic Uncertainties}
\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{ParameterPerturbation.pdf}
\caption{Parameter sweeps using the FAC and PAC show a broad range of control performance in cell 1 ({\bf A}),  in cell 2 ({\bf B}), and in both cells ({\bf C}). Columns show each parameter in the model, rows show the cell which has its parameter perturbed. }
\label{Parameter}
\end{center}
\end{figure*}

Real stochastic processes always have unknown or uncertain mechanisms and parameter values. Although model structures and parameter estimates can often be obtained through fitting to training data, these estimates will never be perfect due to unavoidable measurement errors or limited data sets. 
Even with plentiful and precise training data, model and parameter uncertainties are inevitable due to the fact that even genetically identical single-cells exhibit heterogeneity in parameters due to extrinsic variations.
To asses the control performance under such parameter errors, we performed sensitivity analysis on each model parameter for each cell, then on both cells simultaneously. 
For each parameter, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were examined across a parameter perturbation range from one-tenth to ten-fold of the original values.
Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is varied (all other parameters are fixed), and Fig.\ \ref{Parameter}B shows the control performance as single parameters in Cell 2 is varied. 
Finally, \ref{Parameter}B shows the performance change when a single parameter is changed simultaneously in both Cell 1 and Cell 2 at the same time while keeping other parameters fixed.
In each case the baseline parameters is denoted as a vertical dashed line, and lower values of J (Eq. \ref{XXX}) denote improved performance. 
Solid lines are used to show the performance using the fully aware controller (FAC) that has access to both cells, and the dashed lines show the performance for the partially aware controller (PAC) that only has access to Cell 1.

Modification of parameters were found to produce a broad range of effects. 
For example, increasing $\beta$ in Cell 1 quickly worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Fig.\ \ref{Parameter}A and \ref{Parameter}B). 
In some cases, the effects on performance are not monotonic; for example, increasing $\kappa$ in Cell 1 (Fig.\ \ref{Parameter}A, leftmost column) would be highly advantageous up to a limit after which the control performance degrades rapidly. 
In other cases (such as for the promoter leakage rate, $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=10) perturbations, suggesting that this parameter is not important to control performance.
Figure \ref{Parameter}C shows that even when parameters of both Cell 1 and Cell 2 are jointly changed, these changes could also improve or detract from control performance. 
In particular, the analysis shows that control performance could be improved by modifying the system design either to increase the auto-regulation promoter strength ($\kappa$) or its cooperativity ($\eta$) or to decrease the promoter binding constant ($\beta$) or the protein degradation rate ($\gamma$).

@@

To explore this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.2 and 2.0.  Figure \ref{Volume} (A - F) shows the joint probability distributions (left plots) and marginal probability distributions (right plots) of the system at a low granularity ($\alpha=0.2$), the original granularity ($\alpha=1.0$), and at an increased granularity ($\alpha=2.0$). 
%As the system increases, the scale the size of the dynamics of the system becomes stretched over a larger region in state space. 
%This effect can be observed in the smoothness of C with respect to A despite using the same system and the same controller. 
%\brian[the previous two sentences do not make sense to me.]
Figure \ref{Volume}G shows the trend of the performance versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.  This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate using master equation techniques. To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100. Each SSA was ran for $5\times10^7$ minutes and only the last $4\times10^7$ minutes were sampled to estimate the stationary distribution and calculate the performance score. The performance score estimates of this high granularity SSA using the FAC and PAC were $2.03$ and $5.24$ respectively, which are plotted as dashed lines in Fig.\ \ref{Volume}G. Although it is unclear if further performance improvements could be obtained with further increases to the system scale, for all cases considered so far, we found that both controllers monotonically improved with increased $\alpha$ and that $\mathbf{u}^{\rm FAC}$ always outperforms the $\mathbf{u}^{\rm PAC}$.

@@

However, the improvement in the steady state performance does not come without a cost. Although higher granularity reduces noise and makes it easier to maintain desired states once they have been achieved, noise is necessary to break symmetry between the two cells' dynamics to achieve those states in the first place. This tradeoff is illustrated in Fig \ref{Volume}(H), which plots the control performance over time after the changing the controller goal to exchange the low- and high-target cells with one another.  From the figure, we can see although steady state control performance improves for higher values of $\alpha$, the time taken to reach that steady state performance increases also increases $\alpha$.

@@

Figure \ref{Time} shows the joint distributions (left) and marginal distributions (right) at varying levels of time-delay, with panels A-C showing results for the FAC controller and panels D-F showing results for the PAC controller. Figure \ref{Time}G summarizes these results by plotting the score of both controllers versus the time delay. From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale ($\tau_c = 1 / \gamma = 49 min$) set by the degradation rate of the process (yellow dashed line). At very small time delays ($\tau< 3.4 min$), the FAC outperforms the PAC  but at moderate time delays ($\tau > 3.4 min$) the PAC outperforms the FAC.

\brian[Are these times correct?  Isn't gamma = 0.0203 in Table 1?  tau = 0.07 / 0.0203 (1/min) = 3.44 min. Please check this carefully.]
\michael[I changed tau crit to be 0.1 since 0.07 is arbirarty anyway]
\brian[What are the units in Figure 4G?]
\michael[minutes]

This data show that time delays exceeding $\tau_c = 1 / \gamma = 49$ minutes begins to degrade control performance of both controllers.  \brian[Check these times.] \michael[rewrote the sentece]The choice of best controller depends on the level of time delay in the system. At extreme time delays, both controllers lose their asymmetry, resulting in significantly worse scores ($RMSE = 32.2$ and $23.5$). The control performance of a no-feedback controller was measured which required no inputs Figure\ref{Time}(G) (red line) and achieved a score of $J = 20.0$. These findings indicate that, for large time delays, attempting to improve control with feedback is less effective than using a no-feedback approach.

@@

As discussed in the previous section, increasing the granularity of the system $\alpha$ reduces the randomness to improve the steady state performance but at the cost of slowing down the controller response.  To explore the joint effects of time delays and granularity, Fig,\ \ref{DG}J shows the FAC steady state control performance as a function of the time delay and system granularity $(\tau,\alpha)$ using thirty-two stochastic simulations simulated to steady state at each combination. Control performance errors measured by RMSE improved as $\alpha$ increased when the delays were small ($\tau=1 \ min$ and $\tau=10 \ min$) but not when $\tau=100 \ min$.

\brian[What are the units of tau in fig 5?  This section is confusing in that it is not clear to me how tau and gamma are changing relative to one another when alpha is changed. From Eq 21 it seems that gamma changes with 1/alpha.  If this is correct, does that mean that the critical time changes with $\tau_c = 1/gamma = alpha/gamma ~= 5*alpha$?  If you plot that as a thick dashed dark line in Fig 5J,  would it match to the contours? ] \michael[$tc=1/(0.7*alpha)$ matches the counters well]

Recalling that under the system granularity rescaling (Eq.\ \ref{eq:rescaling}), as $\alpha$ changes, the effective degradation rate scales according to $\gamma' = \gamma/\alpha$.  Therefore, the critical limit for the delays should also change with the system scale according to $\tau_c = 1\alpha/\gamma$. Fig \ref{DG}(J) depicts this critical line and shows that as the time delay approaches and then exceeds this level, the steady state performance become dramatically worse.  \michael[edited this section]

Finally, to understand the model of failure at these longer delays, it is interesting to examine trajectories induced by the controller just above this critical delay. For parameter set denoted by the yellow start in Fig.\ \ref {DG}(J), Fig.\ \ref {DG}(K) shows the controlled response after a long burn in period to achieve steady state. In this case, the application of feedback after a delay leads to a strong oscillatory behavior and worse performance than that achieved without any control at all. This observation further stresses the importance of considering time delays when designing such controllers.

>>>>

While most previous control efforts focussed on deterministic ordinary differential equation (ODE) formulations of the control problem, a few pioneering theoretical works have extended these analysis to include noise and have used stochastic simulations to show how systems can be guided towards unstable points by oscillating around those points \cite{Guarino2020}. 
It has recently been shown that new control techniques that leverage the complete probability distribution information of the system could actually harness the noise of single-cell gene regulation to achieve more complicated control objectives. 
For example, inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that noise could be exploited to achieve independent control of multiple cells using a single input, even despite uncertain parameters or time delays due to maturation of fluorescent proteins or limited observation of the regulatory proteins. 
In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{XXX} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure \ref{cartoons}A, top promoter).
We then proposed the addition of a positive auto-regulation (Figure \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.

@@

\subsection{Model}\label{sec:Model}
To assess the impact of model approximations on the implementation of noise-enhanced control strategies, we begin with the one-species model proposed by May et al. \cite{May2021} (Figure. \ref{cartoons}A). 
This model comprises two reactions for production and degradation of the key protein. 
The nonlinear, UV-dependent production is defined by the following equation:
\begin{equation}
\nu_1(x,t)= \kappa\frac{x^\eta}{x^\eta+\beta^\eta}+ k_0 + u(UV(t)),
\label{prodRate}
\end{equation}
%
%These model dynamics were then simplified to a single ODE with auto-regulation describing the protein accumulation as: 
%\begin{equation}
%\frac{dp}{dt}=u(x_1,x_2)  + \kappa\frac{p^\eta}{p^\eta+\beta^\eta}+ k_0 -\gamma p,
%\label{rateeq}
%\end{equation}
%
where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches its half maximal strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the leakage rate from both promoters; and $u(UV(t))$ is the UV-dependent strength of the T7 promoter. 
Feedback enables the external modulation of the light input using the state of the system, thereby controlling the T7 promoter strength as a function of state rather than time and eliminating the explicit time dependance in $u(UV(t))$.
Protein degradation is assumed to be a first order process with rate $\gamma$:
\begin{equation}
\nu_2(x) = \gamma x.
\end{equation}
All baseline parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$ are presented in Table 1~\cite{May2021} and are fixed throughout the current study.

@@

Optimization of $\mathbf{u}^{FAC}$ was performed using a gradient descent method to minimize $J$. Since the square root is a monotonically increasing function, minimizing $J^2$ results in the same control as would minimizing $J$ directly. Therefore, we calculate the negative gradient $-\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ and walking in a small step $du$ in that direction. The calculation of the gradient is given by
\begin{equation}
\frac{d(J^2)}{d\mathbf{u}^{FAC}}=C \frac{dP}{d\mathbf{u}^{FAC}}
\end{equation}
Where $\frac{dP}{d\mathbf{u}^{FAC}}$ can be solved using general minimized residual calculation of
\begin{equation}
A \frac{dP}{d\mathbf{u}^{FAC}}=B \mathbf{u}^{FAC}
\end{equation}
to find $\frac{dP}{d\mathbf{u}^{FAC}}$.

%Optimization of $\mathbf{u}^{PAC}$ was done by restricting $\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ in ambient space to search a low dimensional subspace with a linear lift operator D. Then the subspace search is given by. 
%\begin{equation}
%\frac{d(J^2)}{d\mathbf{u}^{PAC}}=C \frac{dP}{d\mathbf{u}^{PAC}} = C D \frac{dP}{d\mathbf{u}^{FAC}}  
%\end{equation}

%Root mean squared error was calculated by taking the square root of the score.
%\begin{equation}
%RMSE=\sqrt{J}
%\end{equation}

In \cite{May2021}, controllers were optimized to minimize $J^2$ for a single target state $\mathbf{T} = [10,30]$, but in this work they are extend to different arbitrary target points.

@@

Delays are inherent in any realistic control systems, and it this case delays would be expected to arise due to the time needed for various biochemical reactions such as formation of complete polymerases, activation of promoters, transcription and transport of mature mRNA, and the translation and maturation of protein \cite{Cai2007}. 
Additional delays would also arise from data analysis, decision making, and actuation dynamics. 
To investigate the effects of time delay on control actuation, we devised a simple time-delay stochastic simulation algorithm. 
This algorithm records the state history after each stochastic jump, enabling reconstruction of the population history of the species and the time delayed control input propensity.  
Using this information, the time-delayed control signal at time $t$ can be specified as:

\begin{equation}
u_{\tau}(t)=\left\{
\begin{array}{rl}
      0 ,&\textrm{ for }  t \leq \tau, \\
      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\textrm{ for }   t > \tau,\\
\end{array}\right. 
\label{timeDelaySSA}
\end{equation}
where $\tau$ is the time delay between observation and actuation, ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$), and $x$ gathered from the state history. 
\brian[Explain how you modify the SSA for the delayed reaction SSA.  Look through the literature to cite someone else who has implenented a similar method.]
We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.

@@

\subsection{Dynamic Setpoint Control}\label{sec:Dynamic}
Because optimizing the control law for a static set point as described in Section \ref{sec:Quantification} requires differentiation of the CME (Eq. \ref{CME}) with respect to the control signal at each state, this calculation is approaching the limits of current feasibility.
Extending these calculations to optimize controllers for a dynamically moving set point is much harder and would likely require intractable numerical simulation or development of new mathematical approaches that are beyond the scope of the current study. 
To circumvent this challenge, we instead propose a simple alternative to sweep though a set of controllers over time which target different static points in piecewise fashion to mimic a dynamically moving path.

 We start by describing a discrete path  of target points in state space which cycle at fixed frequency.
  Then, we optimize controllers for each target point and cycling through the controllers over time to create the final controller $\mathbf{R}(\mathbf{x},t)$.
Let $\mathbf{p}(t)$ be a continuous and cyclicly moving point in state space
\begin{equation}
\mathbf{p(t)}=[x_1(t),x_2(t)]
\end{equation}
where $x_1(t)$, and $x_2(t)$ are both cyclic functions of time $\in \mathbf{R}$.
Then, $\mathbf{q}$ is a collection of points $q_i$ which evaluate the cyclic function over its natural period (T) in n steps. 
\begin{equation}
\mathbf{q}=[q_1,\ldots ,q_n]=[\mathbf{p}(t_1),\ldots,\mathbf{p}(t_n)]
\end{equation}
where each $t_i$ comes from linearly spaced points between $0$ and $T$, in $n$ steps 
\begin{equation}
t_i = [t_1,\ldots,t_n]=[0,\frac{1}{n} T,\ldots,\frac{n}{n}T]
\end{equation}
Then, we optimize a set of controllers which target each $q_i$ in the discretized path.
\begin{equation}
\mathbf{U}_i(\mathbf{x})=[\mathbf{u}_1(\mathbf{x}),...,\mathbf{u}_n(\mathbf{x})]
\end{equation}
Next, we cycle through each controller at a fixed frequency by using ($\phi(t)$). 
\begin{equation}
\phi(t)=\lceil \frac{n}{T} mod(t,T)\rceil = \lceil n \ f \ mod(t,\frac{1}{f})\rceil
\end{equation}
where $f$ the cycling frequency given by $1/T$. Then the final control equation is given by 
 \begin{equation}
\mathbf{R}(\mathbf{x},t)=\bf{U}_{\phi(t)}(x)
\end{equation}
Where $\mathbf{R}(\mathbf{x},t)$ returns the control input scalar as a function of state and time. 
The dynamical system equation becomes
\begin{equation}
\frac{d}{dt}\mathbf{P}=\mathbf{A}+\mathbf{B}\mathbf{R}(\mathbf{x},t)
\end{equation}
Since $\mathbf{R}(\mathbf{x},t)$ is a function of time, the system is no longer time invariant. 
\section{Results}

@@

In this section, we seek to show that noise-exploiting controllers are effective at controlling a noisy system to a dynamically moving target point. To do this, we show that control performance is robust to a range of parameter uncertainties without the needing adjustments to the controller. Then, we extend the control performance analysis to understand how time delay and system scale alters control performance. Furthermore, we find regions in target space which are easiest to control by optimizing individual controllers for each target. We extend these analysis to change the target point over time in a Peicewise fashion. 
 \subsection{Stochastic SIMO Optogenetic Control can Remain Effective Despite Small Parameter Errors or Extrinsic Uncertainties}
\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{ParameterPerturbation.pdf}
\caption{Parameter sweeps using the FAC and PAC show a broad range of control performance in cell 1 ({\bf A}),  in cell 2 ({\bf B}), and in both cells ({\bf C}). Columns show each parameter in the model, rows show the cell which has its parameter perturbed. }
\label{Parameter}
\end{center}
\end{figure*}

Real stochastic processes always have unknown or uncertain mechanisms and parameter values. Although model structures and parameter estimates can often be obtained through fitting to training data, these estimates will never be perfect due to unavoidable measurement errors or limited data sets. 
Even with plentiful and precise training data, model and parameter uncertainties are inevitable due to the fact that even genetically identical single-cells exhibit heterogeneity in parameters due to extrinsic variations.
To asses the control performance under such parameter errors, we performed sensitivity analysis on each model parameter for each cell, then on both cells simultaneously. 
For each parameter, $\mathbf{u}^{\rm FAC}$ (solid lines) and $\mathbf{u}^{\rm PAC}$(dashed lines) control performance was examined across a parameter perturbation range from one-tenth to ten-fold of the original values.
Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is varied (all other parameters are fixed), and Figure.\ \ref{Parameter}B shows the control performance as single parameters in Cell 2 is varied. 
Finally, \ref{Parameter}C shows the performance change when a single parameter is changed simultaneously in both Cell 1 and Cell 2 at the same time.

Modification of parameters were found to produce a broad range of effects. 
For example, increasing $\beta$ in Cell 1 quickly worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Figure.\ \ref{Parameter}A and \ref{Parameter}B). 
In some cases, the effects on performance are not monotonic; for example, increasing $\kappa$ in Cell 1 (Figure.\ \ref{Parameter}A, leftmost column) would be highly advantageous up to a limit after which the control performance degrades rapidly. 
In other cases (such as for the promoter leakage rate, $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=10) perturbations, suggesting that this parameter is not important to control performance.
Figure \ref{Parameter}C shows that even when parameters of both Cell 1 and Cell 2 are jointly changed, these changes could also improve or detract from control performance. 
In particular, the analysis shows that control performance could be improved by modifying the system design either to increase the auto-regulation promoter strength ($\kappa$) or its cooperativity ($\eta$) or to decrease the promoter binding constant ($\beta$) or the protein degradation rate ($\gamma$).

@@

To explore this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.2 and 2.0.  Figure \ref{Volume} (A - F) shows the joint probability distributions (left plots) and marginal probability distributions (right plots) of the system at a low granularity ($\alpha=0.2$), the original granularity ($\alpha=1.0$), and at an increased granularity ($\alpha=2.0$). 
%As the system increases, the scale the size of the dynamics of the system becomes stretched over a larger region in state space. 
%This effect can be observed in the smoothness of C with respect to A despite using the same system and the same controller. 
%\brian[the previous two sentences do not make sense to me.]
Figure \ref{Volume}G shows the trend of the performance versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller.  This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate using master equation techniques. To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100. Each SSA was ran for $5\times10^7$ minutes and only the last $4\times10^7$ minutes were sampled to estimate the stationary distribution and calculate the performance score. The performance score estimates of this high granularity SSA using the FAC and PAC were $2.03$ and $5.24$ respectively, which are plotted as dashed lines in Figure.\ \ref{Volume}G. Although it is unclear if further performance improvements could be obtained with further increases to the system scale, for all cases considered so far, we found that both controllers monotonically improved with increased $\alpha$ and that $\mathbf{u}^{\rm FAC}$ always outperforms the $\mathbf{u}^{\rm PAC}$.

@@

However, the improvement in the steady state performance does not come without a cost. Although higher granularity reduces noise and makes it easier to maintain desired states once they have been achieved, noise is necessary to break symmetry between the two cells' dynamics to achieve those states in the first place. This tradeoff is illustrated in Figure\ref{Volume}(H), which plots the control performance over time after the changing the controller goal to exchange the low- and high-target cells with one another.  From the figure, we can see although steady state control performance improves for higher values of $\alpha$, the time taken to reach that steady state performance increases also increases $\alpha$.

@@

Figure \ref{Time} shows the joint distributions (left) and marginal distributions (right) at varying levels of time-delay, with panels A-C showing results for the FAC controller and panels D-F showing results for the PAC controller. Figure \ref{Time}G summarizes these results by plotting the score of both controllers versus the time delay. From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale ($\tau_c = 1 / \gamma = 49$ min) set by the degradation rate of the process (yellow dashed line). At very small time delays ($\tau< 3.4$ min), the FAC outperforms the PAC  but at moderate time delays ($\tau > 3.4$) the PAC outperforms the FAC.

\brian[Are these times correct?  Isn't gamma = 0.0203 in Table 1?  tau = 0.07 / 0.0203 (1/min) = 3.44 min. Please check this carefully.]
\michael[I changed characteristic tau to be 1 since it fits all the curves the best]
\brian[What are the units in Figure 4G?]
\michael[minutes]

These data show that time delays greater than $\tau = 0.1 / \gamma = 4.9$ min begin to degrade FAC control performance and plateau at $\tau_c = 1.0 / \gamma = 49$ min.  \brian[Check these times.] \michael[rewrote the sentence]The choice of best controller depends on the level of time delay in the system. At extreme time delays, both controllers lose their asymmetry, resulting in significantly worse scores ($RMSE = 32$ and $23$). The control performance of simple constant control input without feedback was better than both controlers at very large time delays. Figure\ref{Time}(G) (red line) and achieved a score of $J = 20.0$. These findings indicate that, for large time delays, attempting to improve control with feedback is less effective than using a no-feedback approach.

@@

As discussed in the previous section, increasing the granularity of the system $\alpha$ reduces the randomness to improve the steady state performance but at the cost of slowing down the controller response.  To explore the joint effects of time delays and granularity, Fig,\ \ref{DG}J shows the FAC steady state control performance as a function of the time delay and system granularity $(\tau,\alpha)$ using thirty-two stochastic simulations simulated to steady state at each combination. Control performance errors measured by RMSE improved as $\alpha$ increased when the delays were small ($\tau=1 \ $ min and $\tau=10 \ $ min) but not when $\tau=100 \ $ min.

\brian[What are the units of tau in Figure5?  This section is confusing in that it is not clear to me how tau and gamma are changing relative to one another when alpha is changed. From Eq 21 it seems that gamma changes with 1/alpha.  If this is correct, does that mean that the critical time changes with $\tau_c = 1/gamma = alpha/gamma ~= 5*alpha$?  If you plot that as a thick dashed dark line in Figure5J,  would it match to the contours? ] \michael[$tc=1/(0.7*alpha)$ matches the counters well]

Recalling that under the system granularity rescaling (Eq.\ \ref{eq:rescaling}), as $\alpha$ changes, the effective degradation rate scales according to $\gamma' = \gamma/\alpha$.  Therefore, the critical limit for the delays should also change with the system scale according to $\tau_c = 1\alpha/\gamma$. Figure\ref{DG}(J) depicts this characteristic line and shows that as the time delay approaches and then exceeds this level, the steady state performance become dramatically worse.  \michael[edited this section]

Finally, to understand the model of failure at these longer delays, it is interesting to examine trajectories induced by the controller just above this characteristic delay. For parameter set denoted by the yellow start in Figure.\ \ref {DG}(J), Figure.\ \ref {DG}(K) shows the controlled response after a long burn in period to achieve steady state. In this case, the application of feedback after a delay leads to a strong oscillatory behavior and worse performance than that achieved without any control at all. This observation further stresses the importance of considering time delays when designing such controllers.
